---
title: "Class 11: Structural Bioinformatics pt2."
author: "Ami Pant (A1599358)"
format: html
---

## Background

We saw last day that the PDB has 209,886 entries (Oct/Nov 2025). UniprotKB (i.e. protein sequence database) has 199,579,901 entries.

```{r}
209886/199579901 * 100
```

So the PDB has only 0.1% coverage of the main sequence database.

Enter AlphaFold data base (AFDB). < https://alphafold.ebi.ac.uk/ > that attempts to provide computed models for all sequences in UniProt.

## AlphaFold 

AlphaFOld has three main outputs

- the predicted coordinates (PDB file)
- A local quality score called **pLDDT** (one for each amino acid)
- a second quality score **PAE** predicted aligned error (for each pair of amino-acid)

We can run AlphaFold ourselves if we are not happy with AFDB (i.e. no coverage or poor model).

## Interpreting/analyzing AF results in R

```{r}
results_dir <- "HIVPR_dimer_23119.result/HIVPR_dimer_23119/"
```

```{r}
pdb_files <- list.files(path=results_dir, pattern="*.pdb", full.names = TRUE)
```

```{r}
library(bio3d)
library(bio3dview)

pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")


```



```{r}

rd <- rmsd(pdbs, fit=T)

range(rd)

```

## Analysis

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```





```{r}
pdb <- read.pdb("1hsg")
```


```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, sse=pdb)
points(pdbs$b[2,], typ="l", col="red")
points(pdbs$b[3,], typ="l", col="blue")
points(pdbs$b[4,], typ="l", col="darkgreen")
points(pdbs$b[5,], typ="l", col="orange")
abline(v=100, col="gray")
```

```{r}
core <- core.find(pdbs)

core.inds <- print(core, vol=0.5)

```

```{r}
xyz <- pdbfit(pdbs, core.inds, outpath="corefit_structures")
```

```{r}
rf <- rmsf(xyz)

plotb3(rf, sse=pdb)
abline(v=100, col="gray", ylab="RMSF")
```


## Predicted Alignment Error

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path=results_dir,
                        pattern=".*model.*\\.json",
                        full.names = TRUE)
```

```{r}
pae2 <- read_json(pae_files[2],simplifyVector = TRUE)
pae5 <- read_json(pae_files[5],simplifyVector = TRUE)

attributes(pae2)

```

```{r}
head(pae2$plddt) 
```

```{r}
pae2$max_pae

pae5$max_pae
```

```{r}
plot.dmat(pae2$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)")
```

```{r}
plot.dmat(pae5$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30))
```

```{r}
plot.dmat(pae2$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          grid.col = "black",
          zlim=c(0,30)) #to plot on the same z range

```






