---
title: "Class08: Breast Cancer Analysis Project"
author: "Ami Pant (A1599358)"
format: pdf
toc: TRUE
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. We will extend what we learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. 

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.


## Data import

The data is available as a CSV from the class website:

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)

head(wisc.df)
```

Make sure we do not include sample id or diagnosis columns in the data that we analyze below.

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
wisc.data <- wisc.df[,-1]
dim(wisc.data)
```


## Exploratory data analysis 

> Q1. How many obsevations are in this dataset?

```{r}
dim(wisc.data)
```


There are `r nrow(wisc.data)` observations in this data set.

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

There are 212 observations that have a malignant diagnosis. 

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#colnames(wisc.data)

length(grep("_mean", colnames(wisc.data)))
```

There are 10 variables/features in the data set that are suffixed "_mean".

## Principal Component Analysis

The main function in base R for PCA is called `prcomp()`. An optional argument `scale.` should nearly always be switched to `scale. = T) for this function.

```{r}
wisc.pr <- prcomp(wisc.data, scale = T)
summary(wisc.pr)

```



Let's make our main result figure - the "PC plot" or "score plot", "orrienation plot"

```{r}
library(ggplot2)
ggplot(wisc.pr$x) +aes(PC1, PC2, col = diagnosis) +geom_point()

```


> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

```{r}
wisc_pr_summary <- summary(wisc.pr)
wisc_pr_summary
```

The first principal captures 0.4427 of the variance.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Based on the cumulative proportions of each principal components, we need the first 3 principal components to describe at least 70% of the original variance in the data since PC3 is 0.72636.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

Seven principal components are necessary to describe at least 90% of the original variance in the data. The cumulative proportion of PC7 is 0.91010.

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```

It is very difficult to read and understand this plot so we can not make any conclusions from it. This is a result of the data set being very large.

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}

ggplot(wisc.pr$x) +aes(PC1, PC3, col = diagnosis) +geom_point()

```

Changing the comparison to PC3, brings the points down closer to y = 0. The spread of PC2 is broader (range of y axis 5 to -10) compared to PC3 (range of y axis is 5 to -5). The overall clusters in the PC1 vs PC3 also are less obvious than those in PC1 vs PC2.



> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC.

```{r}
wisc.pr$rotation[,1]
```

The component of the loading vector of concave.points_mean is -0.26085376 in PC1. 

## Heirarchial Clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")
```


> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)
```

The height at which the clustering model has 4 clusters is 19.


```{r}
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)

wisc.hclust <- hclust(data.dist, method = "ward.D2")
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)

wisc.hclust <- hclust(data.dist, method = "single")
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)

wisc.hclust <- hclust(data.dist, method = "average")
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)


```

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

ward.D2 gives the best results for the same data set since it breaks the data into 2 distinct clusters. All the other methods result in dendrograms with really difficult to distinguish the different bigger clusters. 

## Combining PCA and Clustering

```{r}
d <- dist(wisc.pr$x[,1:3])
wisc.pr_hclust <- hclust(d, method = "ward.D2")
#
plot(wisc.pr_hclust)
abline(h = 70, col = "red")
```

## Combining techniques 

Get my cluster membership vector
```{r}
grps <- cutree(wisc.pr_hclust, k =2 )
table(grps)
```


```{r}
plot(wisc.pr_hclust)
abline(h = 35, col = "red")
```

> Q13. How well does the newly created model with four clusters separate out the two diagnoses?

Make a wee "cross-table"
```{r}
table(grps, diagnosis)
```

The clustering results in 179 true positives, 24 false positives, 333 true negatives and 33 false negatives.

Sensitivity: TP/(TP+FN) = 179/(179+33) = 0.8443
Specificity: TN/(TN + FN) = 333/(333+33) = 0.9098

```{r}
grps <- cutree(wisc.pr_hclust, k =4 )
table(grps)
table(grps, diagnosis)
```


The newly created model with four clusters, doesn't separates the two diagnoses super well. The first and fourth clusters separate the diagnosis pretty well with cluster one being 100% true positives malignant patients and cluster 4 being 149 out of 150 being true negative. Cluster 2 and 3 has a significant mix of both diagnoses without good separation. 


> Q14. How well do the hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}

hclust.single <- hclust(data.dist, method = "single")
grps <- cutree(hclust.single, k =4 )
table(grps)
table(grps, diagnosis)


hclust.average <-  hclust(data.dist, method = "average")
grps <- cutree(hclust.average, k =4 )
table(grps)
table(grps, diagnosis)

hclust.complete <- hclust(data.dist, method = "complete")
grps <- cutree(hclust.complete, k =4 )
table(grps)
table(grps, diagnosis)

```

Single and average hclust hierarchical clustering techniques are definitely not better than the ward.D2 clustering method because most of the patients are clustered into one (the first) cluster. The clustering method of complete seems slightly better than single and average, however the two main clusters (1 and 3) include a mixture of patients with benign and malignant tumors in them. There are not as many patients included in cluster 2 and 4, meaning not all 4 clusters are super relevant. 



> Q16. Which of these new patients should we prioritize for follow up based on your results?

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
grps <- cutree(wisc.pr_hclust, k = 2)
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

Based on the result, we should prioritize patient 2 for a follow up because they fall within the malignant cluster. 







