---
title: "Class05: Data Viz with ggplot"
author: "Ami Pant (A15991358)"
format: pdf
---

Today we are playing with plotting and graphics in R. 

There are lots of ways to make cool figures in R.
There is "base" R graphics(`plot()`, `hist()`, `boxplot()` etc.)

There is also add-on packages, like **ggplot**.

```{r}
head(cars)
```

Let's plot this with "base" R:
```{r}
plot(cars)
```

```{r}
head(mtcars)
```

Let't plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch = 10)
```

```{r}
hist(mtcars$mpg)
```

## GGPLOT

The main function of the ggplot2 package is `ggplot()`. First I need to install the **ggplot2** package. I can install any package with the function `install.packages()`. 

> **Note** I never want to run 'install.packages()` in  my quarto sourse document!!


```{r}
library(ggplot2)

ggplot(cars) +
  aes(speed, dist) +
  geom_point()
```


```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least three things: 

- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes()`)
- The **geom** (given by `geom_point()`)

> For simple canned graphs "base" R is nearly always faster 

### Adding more layers

Let's add a line 

```{r}
ggplot(cars) +
  aes(speed, dist) +
  geom_point() +
  geom_line()
```

Let's add a line, subtitle, and caption as well as axis labels
```{r}
ggplot(cars) +
  aes(speed, dist) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Silly PLot", x = "Speed (mph)", y = "Distance (ft)", caption = "when is tea time anyway??") +
  theme_bw()
```


## Plot some expression data

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this wee dataset? 

There are `r nrow(genes)` in this data set. 

> Q2. How many "up" regulated genes are there

```{r}
sum(genes$State == "up")

```

There are `r sum(genes$State == "up")` UP genes.

```{r}
table(genes$State)
```

```{r}
p <- ggplot(genes, aes(Condition1, Condition2, color = State)) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red"))
```


```{r}
p +labs(title = "look at me", )
```

Silly examples of adding labels 

```{r}
library(ggrepel)
ggplot(genes, aes(Condition1, Condition2, 
                  color = State, 
                  label = Gene)) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_text_repel(max.overlaps = 100) +
  theme_bw()
```

## Going Further

Playing with some different layers and the gapminder dataset ...

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```

```{r}
tail(gapminder)
```

A first plot 

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent, size = pop)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~continent) +
  theme_bw()

```


```{r}
library(patchwork)

# Setup some example plots 
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))

# Use patchwork to combine them here:
(p1 | p2) / (p3 | p4)

```


