---
title: "Class 17: RNASeq Pop Scale Analysis"
author: "Ami Pant (A15991358)"
format: html
---

## Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their 
corresponding median expression levels for each of these genotypes. 


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
table(expr$geno)
```
```{r}
median(expr$exp[expr$geno == "A/A"])

median(expr$exp[expr$geno == "A/G"])

median(expr$exp[expr$geno == "G/G"])


```

The sample size for the A/A is 108, A/G is 233, and G/G is 121. The median expression levels for A/A is 31.24847, A/G is 25.06486, and G/G is 20.07363.   

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative 
expression value between A/A and G/G displayed in this plot? Does the SNP effect the 
expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) +aes(geno, exp, fill = geno) +geom_boxplot(notch = T)
```

Based on the histogram, this specific SNP does have an effect on the expression of ORMDL3. In the presence of homozygous allele SNPs there is lower gene expression of ORMDL3 than the wildtype (A|A) and heterozygous alleles. The heterozygous allele (A|G) also result in a lowered expression of ORMDL3, but not as low as the homozygous SNP (G|G)
