---
title: "Class 16: Obtaining and processing SRA datasets on AWS"
author: "Ami Pant (A15991358)"
format: html
---

## Working with RNA-Seq data

> Q1. How would you check that these files with extension ‘.fastq’ actually look like what we expect for a FASTQ file? You could try printing the first few lines to the shell standard output:

On AWS

$ head *.fastq

> Q2. How could you check the number of sequences in each file?

On AWS

$ grep -c "@SRR" *.fastq

> Q3. Check your answer with the bottom of the file using the tail command and also check the matching mate pair FASTQ file. Do these numbers match? If so why or why not?

On AWS

$ ls *.fastq

> Q4. Can you run kallisto to print out it’s citation information?

On AWS

$ kallisto --cite

> Q5. Have a look at the TSV format versions of these files to understand their structure. What do you notice about these files contents?

On AWS

$ head \*_quant/*.tsv

## Downstream analysis

```{r}
library(tximport)
```


```{r}
folders <- dir(pattern = "SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

```{r}
colSums(txi.kallisto$counts)
```

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```


## Principal Component Analysis

```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q6. Use ggplot to make a similar figure of PC1 vs PC2 and a seperate figure PC1 vs PC3 and PC2 vs PC3.

```{r}
library(ggplot2)
library(ggrepel)
```

```{r}
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

```{r}
ggplot(y) +
  aes(PC1, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

```{r}
ggplot(y) +
  aes(PC2, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```




