---
title: "class10: Halloween Candy Miniproject"
author: "Ami Pant (A1599538)"
format: pdf
---

```{r, message = FALSE}
library(dplyr)
library(skimr)
library(ggplot2)
library(ggrepel)
library(corrplot)
```



```{r}
candy_file <- "candy-data.csv"

candy <- read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 different candies in this dataset.


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruity candy types in the dataset.


> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}

candy["Reese's Peanut Butter cup",]$winpercent

```

My favorite candy is Reese's peanut butter cups, which has a 84.18 win percent.

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

Kitkat has a 76.76% win percent.

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

Tootsie Roll Snack Bars have a 49.65% win percent.


```{r}
candy %>% select(winpercent)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r, error = T}
skim(candy)
```

The winpercent variable looks like it is on a different scale to the majority of the dataset.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A 1 represents the specific candy is chocolate based and a 0 means it is not.


> Q8. Plot a histogram of winpercent values

```{r}
ggplot(candy) +aes(winpercent) +geom_histogram(bins = 20)
```

> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) + aes(winpercent) +geom_density()
```

The distribution is not symmetric. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

The mean is 50.32 which is above 50% but the median is 47.83 which is below 50%. We trust median more as a measure of centrality when there are outliers.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset

choc.inds <- as.logical(candy$chocolate)
choc.inds

choc.candy <- candy[choc.inds,]
choc.candy

# 2. Extract their `winpercent` values

choc.win <- choc.candy$winpercent
choc.win

# 3. Find the mean of these values

choc.mean <- mean(choc.win)
choc.mean

# 4-6. Do the same for the fruity candy
fruit.inds <- as.logical(candy$fruity)
fruit.inds

fruit.candy <- candy[fruit.inds,]
fruit.candy

fruit.win <- fruit.candy$winpercent
fruit.win

fruit.mean <- mean(fruit.win)
fruit.mean

# 7. Which mean value is higher
 
choc.mean > fruit.mean

```

The mean chocolate candy rankings is higher than the ranking of fruity candy. 


> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

Based on two sample t-test, the difference between chocolate and fruity candy is statistically significant.


> Q13. What are the five least liked candy types in this set?

```{r}
candy %>% arrange(winpercent) %>% head(5)
```

The five least liked candies are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. I prefer using piping to answer this question becasue I find it easier to read the code.

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

The top 5 candies are Reese's Peanut Butter cup, Reese's Minatures, Twix, Kit Kat, and Snickers.

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <-  "chocolate"
my_cols[as.logical(candy$bar)] <-  "brown"
my_cols[as.logical(candy$fruity)] <-  "pink"



ggplot(candy) +aes(x = winpercent, y = reorder(rownames(candy), winpercent)) +geom_col(fill = my_cols)
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy, based on the bar graph, is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

## Winpercent vs Pricepercent


```{r}
my_cols[as.logical(candy$fruity)] <-  "blue"
ggplot(candy) +aes(winpercent, pricepercent, label = rownames(candy)) +geom_point(col = my_cols) +geom_text_repel(col = my_cols)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Fruity candy tends to be the highest rank for the least money.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy %>% arrange(desc(pricepercent)) %>% head(5) %>% arrange(winpercent)
```

The top five most expensive candy types in the dataset are Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular among these five is Nik L Nip.

## Correlation

```{r}
cij <- cor(candy)
```

```{r}
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruit and chocolate tend to be anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar as well as chocolate and winpercent to be positively correlated.

## Principal Component Analysis

The main function in base R for this is `prcomp()` and we want to set `scale=TRUE` here:

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

Let's look at our first main result figure - the "PC plot" on PC1 vs PC2

```{r}
ggplot(pca$x) +aes(PC1, PC2, label = rownames(pca$x)) +geom_point(col = my_cols) +geom_text_repel(col = my_cols)

```

Don't forget about your variable "loadings" - how the original variables contribute to your new PCs

```{r}
ggplot(pca$rotation) +aes(PC1, rownames(pca$rotation)) +geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Pluribus, hardess, and fruity-ness are strongly picked up by PC1 in a positive direction. This does make sense since most fruity candies tend to be hard candies. 




